---
title: "Week 1 Handout: Computer Test"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "January 13, 2017"
output: pdf_document
geometry: margin=2.54cm
---

The goal of this short assignment to to make sure that your personal computing environment is functioning as needed for the course.
From the RMarkdown file (`Week1_CompTest.Rmd`), you will print a simple statement and then retrieve information about the operating system and versions of software on your computer.
You will then  **Knit** the text and code into a PDF file.
Last, you will submit this as your first assignment, which is due on **January 18^th^, 2017 at 12:00 PM (noon)**.

1. Open `Week1_CompTest.Rmd` in RStudio (see Week1 folder)

2. Edit your name in the *author* line above (line 3)

3. Print a message from R using the `print` command (practice in the console):

    Ex. Print "*Hello, World!*"
        
```{r}
print("Hello, World!")
```

4. Now, print your own statement (it can be anything):
```{r}
    
    
```

5. Now run the code below to retrieve information about your local computing environment. 
You do not need to make any changes to this code. 
The output below is what was returned from Jay's computer:

```{r}
info <- unname(c(Sys.info()[6],
          .Platform$OS.type,
          Sys.info()[1:2],
          R.version.string,
          .Platform$GUI,
          system("git --version", intern = TRUE)))
info
```

