---
title: "RStudio Assignment: R Test"
author: "Student Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

## Directions:
This is an RMarkdown file.
It will allow you to integrate text and R code into one file.
Please complete parts 1 and 2 below.
Part 3 should be left alone (it should work as needed)
When you are done, **Knit** the text and code into a PDF file.
Basically, just press the **Knit** button above.
After Knitting, please submit via GitHub. 
Review your in-class Git assignment if needed.
It is due on **January 18^th^, 2015 at Noon**.  

## Assignment:

1. Edit your name in the *author* line above (line 3)

2. Print a message in the R terminal

    Ex. Print "*Hello, World!*"
```{r}


```

    Now, print your own statement
```{r}


```

3. Tell us about your system. Basically, just leave this part alone and it will do magic

**Do Not Modify This Part**

```{r}
info <- unname(c(Sys.info()[6],
          .Platform$OS.type,
          Sys.info()[1:2],
          R.version.string,
          .Platform$GUI,
          system("git --version", intern = TRUE)))
info
```

